Subset coercions in \Coq \\ \small{Draft - \today}

### 2006-08-23 papers/article/main.tex
Bib fixes
British spell checking, e.g. and i.e. corrections

\abstract{We propose a new language for writing programs with dependent
  types which can be elaborated into partial \Coq terms. This language permits
  to establish a phase distinction between writing and proving
  algorithms in the \Coq{} environment.  Concretely, this means allowing
  to write algorithms as easily as in a practical functional programming
  language whilst giving them as rich a specification as desired and
  proving that the code meets the specification using the whole \Coq{}
  proof apparatus. This is achieved by extending conversion to an
  equivalence which relates types and subsets based on them, a technique
  originating from the ``\emph{Predicate subtyping}'' feature of \PVS{}
  and following mathematical convention. The typing judgments can be
  translated to the \CICfull (\CIC{}) by means of an interpretation
  which inserts coercions at the appropriate places. These coercions can
  contain existential variables representing the propositional parts of
  the final term. A prototype implementation of this process has been
  implemented.}

@@

The specification shows that we're defining a partial function and
shows the relation betwenn input and ouput, using a dependent type, yet the
code remains as simple as the \ML{} definition. This is only possible
because we leave the proofs in both branches. After type-checking,
there is an automatic elaboration into partial \Coq terms, which collects
obligations the user has to prove. In our example,
the proof assistant will ask the user to prove that:
\begin{enumerate}
\item The list $tl$ has the property \verb|exists a, a :: tl = hd :: tl|
  and
\item In the context where
  \id{l} is a non-empty list and $l = []$, the list $[]$ has the property
  \verb|exists a, a :: [] = []| (which should be obvious as the context is contradictory).
\end{enumerate}

This solution also provides % PVS:Language,
facilities to express properties with a more mathematical flavor using subsets,
bridging a gap between mathematical convention and type theory.

@@

The remainder of the article is organized as follows: in section \ref{sec:Russell} we present a
type system based on \CIC, which integrates subset ``subtyping'', and
we show how it relates to \CIC{} by means of translations between
judgments of the two systems in section \ref{sec:Translation}. Then, we present a prototype
implementation in section \ref{sec:Program}, and finally we give our
conclusions on this work and discuss future directions in section \ref{sec:Conclusion}.

@@

Informally, this idea is already present in mathematics. When you have a
subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, e.g: $f : \{ x : `N `| P \} "->"
X$, you usually prove \emph{first} that you apply it only to elements 
having the desired property. For example $(f~2)$ is a correct application
only if $(P~2)$ is provable.
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. In \Coq{}, we are forced
to apply $f$ to an object $(\elts{2}{p})$ where $p$ is a proof of
$(P~2)$ (figure \ref{fig:cci-subset} presents the definition of the subset type in \Coq).
We would like to be able to prove that our usage of partial functions 
is correct only after the program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and the
properties which we use to reason about them.

@@

\subsection{From \ps{} to subset equivalence}
The \ps{} mechanism \cite{Shankar&Owre:WADT99} %Rushby96:TSE
is an extension of the \PVS type system which internalizes this idea.
Concretely, this means that the following rules are derivable in \PVS:
\begin{center}
  \AXC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \UIC{$`G \vdash t : T$}
  \DP
  \AXC{$`G \vdash t : T$}
  \AXC{$`G \vdash P[t]$}
  \BIC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \DP
\end{center}

The first one formalizes the fact that an object of a subset based on
$T$ is an object of type $T$. The second one permits using an object of
type $T$ as an object of type $\{ x : T `| P[x] \}$, but it generates a
\emph{type checking condition} $`G \vdash P[t]$ which will need to be
discharged later. Effectively, the typing algorithm of \PVS{} collects
the conditions that must be satisfied for the term to be accepted as a
valid definition. However, the acceptance criteria can be rather large in
\PVS{}. By design, when proving in \PVS{}, the trusted code base (TCB
hereafter) is the entire system, not only the typing system but also the
various decision procedures and tactics used to build proofs
automatically or interactively. It is nonetheless a widely used proof
assistant and the predicate subtyping feature has apparently helped to
build a consequent library of certified code. On the other hand, \Coq
has a small TCB and greater expressiveness but less automation and
methodology to build certified programs. We capitalize on the \PVS
success to make \Coq more usable for this kind of tasks, and as we will see,
it will have other benefits.

\subsection{A weaker type system}
To formalize this idea in \Coq, we simply weaken the type system so that
it doesn't require the terms to contain the proof components for objects
of subset types. This permits to have a simple language for code while
retaining the richness of \Coq's specification language. Once we have a
derivation in this new type system, we can translate it to a partial
\Coq derivation, where the missing parts are represented by
metavariables. It can then be completed by instantiating these holes
with actual proofs.

@@

    \vspace{\infvspace}
    \Conv\DP
  \end{center}
  \vspace{-2em}
  \caption{\CIC typing judgment}
  \label{fig:cic-decl-rules}
  \vspace{-1em}
\end{figure}

@@

%
\Russell{}'s type system is based on the Calculus of (Co-)Inductive
Constructions (figure \vref{fig:cic-decl-rules}) \cite{coquandhuet88},
with sigma types but without universes.  This restriction may be removed
in future work, but cause no problem for our main purpose which is
programming. We omit (co-)inductive constructs here, as they can be
considered as constants and leave their treatment as future work. The
judgment $`G \typed t : T$ means $t$ is a well-typed term of type $T$
in environment $`G$.  In \Russell, the conversion rule is replaced by a
new subsumption rule \irule{Subsum} (figure
\vref{fig:typing-decl-rules}) which will implement the subset
equivalence. The judgment $`G \typed T \subd U : s$ means $T$ is
equivalent to $U$ in environment $`G$, both sorted with $s$.
Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $(s_1, s_2, s_2)$. We allow products $\Pi x :
A, B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ \_ : unit `| A \}, B$ products to benefit from
the subtyping mechanism.

@@

Likewise, we need to restrict the system so that it becomes property
irrelevant for the propositions appearing in subset types. Indeed, in
\PVS{} you can't manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgment $0 : \mysubset{x}{`N}{x \neq 0}$, but
in a classical Calculus of Constructions with sigma types or inductive
types, the subset type is a first class object defined with proper
introduction and elimination terms (figure \vref{fig:cci-subset}). What
we need to do is forbid the use of these terms in \Russell{} terms,
leaving only the type former (figure \vref{fig:typing-decl-rules}).

@@

\subsubsection{Equivalence}
We have renamed the technique from \ps{} to 
\emph{subset equivalence} because we have a symmetric relation, contrary
to usual subtyping relations. It also conveys the idea that it can
include the usual $\beta\pi$-conversion directly in the judgment.
The judgment $`G \typed T \subd{} U : s$ (figure
\vref{fig:conversion-decl-rules}) reads $T$ is equivalent to $U$ in
environment $`G$, both being sorted by $s$.

@@

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgment. We use a judgmental equality here, which will be refined by
the usual conversion relation later. The \irule{SubSym} and
\irule{SubTrans} rules ensure that our judgment constructs an \emph{equivalence} and has
proper modularity. The next two rules (\irule{SubProd} and
\irule{SubSigma}) do context closure for dependent
products and sums. It is remarkable that we use contravariance for
% \cite{journals/toplas/Castagna95}
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgments because we have symmetry. However, it will
become important when we create coercions (see figure
\vref{fig:coerce-impl-rules-short}).

@@

\subsubsection{Properties}
We have proved the metatheory for this system in \Coq{}
\cite{sozeau:Coq/Russell/meta-web}. We have assumed that it is strongly
normalizing (SN) but we have proved subject reduction (SR) for it: 
if $`G \typed t : T$ and $t "->"_{\beta\pi} t'$ then $`G \typed t' :
T$ (our proof of SR doesn't depend on SN).
Gang Chen \cite{ChenPhD} has studied various type systems from the
$\lambda$-cube extended by subtyping or coercive subtyping
\cite{DBLP:conf/csl/Luo96}, including the Calculus of Constructions, and
proved such results as SN and SR for them. It requires a very careful
analysis of the system to avoid cycles in the proof due to the presence
of dependent types and conversion. However we preferred to adapt the method of
Robin Adams \cite{adams:PTSEQ} to prove SR, because Gang Chen's method
seemed very tied to the peculiarities of the system he studied. 
Strong normalization could also be proved, perhaps more easily,
using the generic proof method developed in \cite{geuvers95}, by finding
a proper interpretation of \Russell terms into the Calculus of
Constructions.

So, apart from strong normalization, we have shown all the usual structural
and metatheoretic properties of a dependent programming language, like
weakening, thinning and substitution, stability by context coercion etc\ldots
We stop here on the subject of theoretical properties of this first system, as
it gives no new insights for our purpose and we will focus more on the
algorithmic system's properties. Besides, the details of the
formalization and the proof fall outside the scope of this paper.
% The essential properties of the $\subd$ judgement are 
% its reflexivity: if $`G \typed A : s$, then $`G \typed A \subd A$,
% symetry: if $`G \typed A, B : s$ then $`G \typed A \subd B "<=>" `G \typed
% B \subd A$ and transitivity by \irule{SubTrans}.

@@

\begin{paragraph}{}
We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
%We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
%$T  `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
Our algorithmic judgment $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. 
% This notion is presented in \PVS{} by
% means of a $\mu_0$ operator which was readily adaptable in our setting
% (figure \vref{fig:mualgo-definition}). In Gang Chen's work, the notion is formalized by a
% new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator
% $\mualgo$, except ours finds $\Sigma_{lub}$s too.
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}
    \def\fCenter{\typea}
    \begin{center}
%     \WfEmpty\DP

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgment for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity and symmetry rules which are not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Admissibility of transitivity]
  If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
  \suba V : s$.
\end{theorem}
\begin{proof}
  By induction on the sum of the depths of the two derivations.
\end{proof}
\end{paragraph}

@@

Our interpretation will build a full-fledged \CIC term from its
algorithmic skeleton and a rich type written in \Russell. Obviously, we
cannot infer proof terms where they are needed in \CIC, but we can build
a \emph{partial} term, leaving typed holes where proofs are
needed. Hence we add a rule to build existential variables (or
metavariables) in the target calculus ($\typeq$ denotes the new system's typing judgment):

@@

The proof of this proposition is the main technical contribution of this
paper. It also represented the main difficulty in our work. 
The remainder of this section is organized as follows: first we will
define the explicit coercion derivation algorithm which will permit to
put proof obligations in the terms, then we will define the
interpretation of terms which is mutually recursive with the precedent
algorithm. Finally we will present a proof sketch of the aforementioned
proposition.

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations, hence we can derive a function from
the judgment which build a coercion given two types (we denote it $\coerce{`G}{T}{U}$). 
The side-conditions of the rules have not changed, so we omit them for
improved readability. The judgment $\subimpl{`G}{c}{T}{U}$ builds a
coercion $c$ from $T$ to $U$ in context $`G$, that is an object which
coerces an object of type $\ip{T}{`G}$ to an object of type $\ip{U}{`G}$
in context $\ipG{`G}$. We use the algorithmic types to drive the
derivation but the resulting object will be a well-typed \Coq{} term. We
could also directly use the interpreted types but that complicates the
proof.

@@

Coercions are formalized as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instantiation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(ie, it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied.
Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgment $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and adapted for checking coherence 
of coercions, which we don't need here as we can get unicity
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respects some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this wasn't adapted to our setting, so we invented this
new presentation.

@@

We first prove reflexivity, symmetry and transitivity of the coercion
derivation algorithm, which extends previous proofs for the algorithmic
system with properties on
the generated coercions. Then we show substitutivity of the
interpretation: if $`G, x : U, `D \type t : T$ then $\ip{t[u/x]}{`G,
 `D[u/x]} = \ip{t}{`G, x : U, `D}[\ip{u}{`G}/x]$. We extend the notion
of coercion to contexts next and show stability of judgments under
context coercion. It is then possible to prove a lemma about commutativity of
substitutions with the interpretation and coercions. Then we get an
important corollary: if $`G \type u : U$, $\subimpl{`G}{c}{U}{V}{s}$ and
$`G, x : V \type T : s$ then $\ip{T[u/x]}{`G} `= \ip{T}{`G, x :
  V}[c[\ip{u}{`G}]/x]$. Essentially, it shows that coercions inserted
by the interpretation depend only on the context and hence could be
added later using substitution instead. This allows us to show that
equivalence is conserved by interpretation and finally that our
interpretation is correct.

\section{The \Program{} tactic}
\label{sec:Program}
This mechanism has been implemented in \Coq and tested on simple
examples. We actually generate coercions simultaneously with typing 
in the implementation, which is sound thanks to the previous proofs. Our
typing is a clone of \Coq's original typing algorithm, so we benefit
from all the features of \Coq including implicit variables, notations and the existing
coercion system. The type-checker also transforms pattern-matching so
that an equality between the matched term and the pattern is present in
each branch typing context and it allows the user to put explicit holes in the term, for example:
\begin{code}
  Definition head \{ l : list A | l <> 0 \} : A :=
    match l with
    | hd :: tl -> hd
    | [] -> _
    end.
\end{code}
Here the second branch can't be taken because we would have $l = []$ and
$l \neq 0$, but we may not have any placeholder element of type $A$.

@@

\paragraph{Acknowledgments}
I'd like to thank Christine Paulin-Mohring for directing this work.
\bibliography{my,bib-joehurd,barras,pvs-bib,bcp,Luo,biblio-subset,%
cparent,demons,demons2,demons3}
\bibliographystyle{splncs}
% \renewcommand{\thefootnote}{}
% \footnotetext{This article was typeset in \LaTeX~using the
%   \texttt{Concrete Math} font}

>>>>

\abstract{We propose a new language for writing programs with dependent
  types which can be elaborated into partial \Coq terms. This language permits
  to establish a phase distinction between writing and proving
  algorithms in the \Coq{} environment.  Concretely, this means allowing
  to write algorithms as easily as in a practical functional programming
  language whilst giving them as rich a specification as desired and
  proving that the code meets the specification using the whole \Coq{}
  proof apparatus. This is achieved by extending conversion to an
  equivalence which relates types and subsets based on them, a technique
  originating from the ``\emph{Predicate subtyping}'' feature of \PVS{}
  and following mathematical convention. The typing judgements can be
  translated to the \CICfull (\CIC{}) by means of an interpretation
  which inserts coercions at the appropriate places. These coercions can
  contain existential variables representing the propositional parts of
  the final term. A prototype implementation of this process has been
  implemented.}

@@

The specification shows that we're defining a partial function and
shows the relation between input and output, using a dependent type, yet the
code remains as simple as the \ML{} definition. This is only possible
because we leave the proofs in both branches. After type-checking,
there is an automatic elaboration into partial \Coq terms, which collects
obligations the user has to prove. In our example,
the proof assistant will ask the user to prove that:
\begin{enumerate}
\item The list $tl$ has the property \verb|exists a, a :: tl = hd :: tl|
  and
\item In the context where
  \id{l} is a non-empty list and $l = []$, the list $[]$ has the property
  \verb|exists a, a :: [] = []| (which should be obvious as the context is contradictory).
\end{enumerate}

This solution also provides % PVS:Language,
facilities to express properties with a more mathematical flavour using subsets,
bridging a gap between mathematical convention and type theory.

@@

The remainder of the article is organised as follows: in section \ref{sec:Russell} we present a
type system based on \CIC, which integrates subset ``subtyping'', and
we show how it relates to \CIC{} by means of translations between
judgements of the two systems in section \ref{sec:Translation}. Then, we present a prototype
implementation in section \ref{sec:Program}, and finally we give our
conclusions on this work and discuss future directions in section \ref{sec:Conclusion}.

@@

Informally, this idea is already present in mathematics. When you have a
subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, say $f : \{ x : `N `| P \} "->"
X$, you usually prove \emph{first} that you apply it only to elements 
having the desired property. For example $(f~2)$ is a correct application
only if $(P~2)$ is provable.
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. In \Coq{}, we are forced
to apply $f$ to an object $(\elts{2}{p})$ where $p$ is a proof of
$(P~2)$ (figure \ref{fig:cci-subset} presents the definition of the subset type in \Coq).
We would like to be able to prove that our usage of partial functions 
is correct only after the program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and the
properties which we use to reason about them.

@@

\subsection{From \ps{} to subset equivalence}
The \ps{} mechanism \cite{Shankar&Owre:WADT99} %Rushby96:TSE
is an extension of the \PVS type system which internalises this idea.
Concretely, this means that the following rules are derivable in \PVS:
\begin{center}
  \AXC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \UIC{$`G \vdash t : T$}
  \DP
  \AXC{$`G \vdash t : T$}
  \AXC{$`G \vdash P[t]$}
  \BIC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \DP
\end{center}

The first one formalises the fact that an object of a subset based on
$T$ is an object of type $T$. The second one permits using an object of
type $T$ as an object of type $\{ x : T `| P[x] \}$, but it generates a
\emph{type checking condition} $`G \vdash P[t]$ which will need to be
discharged later. Effectively, the typing algorithm of \PVS{} collects
the conditions that must be satisfied for the term to be accepted as a
valid definition. However, the acceptance criteria can be rather large in
\PVS{}. By design, when proving in \PVS{}, the trusted code base (TCB
hereafter) is the entire system, not only the typing system but also the
various decision procedures and tactics used to build proofs
automatically or interactively. It is nonetheless a widely used proof
assistant and the predicate subtyping feature has apparently helped to
build a consequent library of certified code. On the other hand, \Coq
has a small TCB and greater expressiveness but less automation and
methodology to build certified programs. We capitalise on the \PVS
success to make \Coq more usable for this kind of tasks, and as we will see,
it will have other benefits.

\subsection{A weaker type system}
To formalise this idea in \Coq, we simply weaken the type system so that
it doesn't require the terms to contain the proof components for objects
of subset types. This permits to have a simple language for code while
retaining the richness of \Coq's specification language. Once we have a
derivation in this new type system, we can translate it to a partial
\Coq derivation, where the missing parts are represented by
metavariables. It can then be completed by instantiating these holes
with actual proofs.

@@

    \vspace{\infvspace}
    \Conv\DP
  \end{center}
  \vspace{-2em}
  \caption{\CIC typing judgement}
  \label{fig:cic-decl-rules}
  \vspace{-1em}
\end{figure}

@@

%
\Russell{}'s type system is based on the Calculus of (Co-)Inductive
Constructions (figure \vref{fig:cic-decl-rules}) \cite{coquandhuet88},
with sigma types but without universes.  This restriction may be removed
in future work, but cause no problem for our main purpose which is
programming. We omit (co-)inductive constructs here, as they can be
considered as constants and leave their treatment as future work. The
judgement $`G \typed t : T$ means $t$ is a well-typed term of type $T$
in environment $`G$.  In \Russell, the conversion rule is replaced by a
new subsumption rule \irule{Subsum} (figure
\vref{fig:typing-decl-rules}) which will implement the subset
equivalence. The judgement $`G \typed T \subd U : s$ means $T$ is
equivalent to $U$ in environment $`G$, both sorted with $s$.
Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $(s_1, s_2, s_2)$. We allow products $\Pi x :
A, B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ \_ : unit `| A \}, B$ products to benefit from
the subtyping mechanism.

@@

Likewise, we need to restrict the system so that it becomes property
irrelevant for the propositions appearing in subset types. Indeed, in
\PVS{} you can't manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgement $0 : \mysubset{x}{`N}{x \neq 0}$, but
in a classical Calculus of Constructions with sigma types or inductive
types, the subset type is a first class object defined with proper
introduction and elimination terms (figure \vref{fig:cci-subset}). What
we need to do is forbid the use of these terms in \Russell{} terms,
leaving only the type former (figure \vref{fig:typing-decl-rules}).

@@

\subsubsection{Equivalence}
We have renamed the technique from \ps{} to 
\emph{subset equivalence} because we have a symmetric relation, contrary
to usual subtyping relations. It also conveys the idea that it can
include the usual $\beta\pi$-conversion directly in the judgement.
The judgement $`G \typed T \subd{} U : s$ (figure
\vref{fig:conversion-decl-rules}) reads $T$ is equivalent to $U$ in
environment $`G$, both being sorted by $s$.

@@

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgement. We use a judgemental equality here, which will be refined by
the usual conversion relation later. The \irule{SubSym} and
\irule{SubTrans} rules ensure that our judgement constructs an \emph{equivalence} and has
proper modularity. The next two rules (\irule{SubProd} and
\irule{SubSigma}) do context closure for dependent
products and sums. It is remarkable that we use contravariance for
% \cite{journals/toplas/Castagna95}
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symmetry. However, it will
become important when we create coercions (see figure
\vref{fig:coerce-impl-rules-short}).

@@

\subsubsection{Properties}
We have proved the metatheory for this system in \Coq{}
(\url{http://www.lri.fr/~sozeau/research/russell/proof.en.html}).
 We have assumed that it is strongly
normalising (SN) but we have proved subject reduction (SR) for it: 
if $`G \typed t : T$ and $t "->"_{\beta\pi} t'$ then $`G \typed t' :
T$ (our proof of SR doesn't depend on SN).
Gang Chen \cite{ChenPhD} has studied various type systems from the
$\lambda$-cube extended by subtyping or coercive subtyping
\cite{DBLP:conf/csl/Luo96}, including the Calculus of Constructions, and
proved such results as SN and SR for them. It requires a very careful
analysis of the system to avoid cycles in the proof due to the presence
of dependent types and conversion. However we preferred to adapt the method of
Robin Adams \cite{adams:PTSEQ} to prove SR, because Gang Chen's method
seemed very tied to the peculiarities of the system he studied. 
Strong normalisation could also be proved, perhaps more easily,
using the generic proof method developed in \cite{geuvers95}, by finding
a proper interpretation of \Russell terms into the Calculus of
Constructions.

So, apart from strong normalisation, we have shown all the usual structural
and metatheoretic properties of a dependent programming language, like
weakening, thinning and substitution, stability by context coercion etc\ldots
We stop here on the subject of theoretical properties of this first system, as
it gives no new insights for our purpose and we will focus more on the
algorithmic system's properties. Besides, the details of the
formalisation and the proof fall outside the scope of this paper.
% The essential properties of the $\subd$ judgement are 
% its reflexivity: if $`G \typed A : s$, then $`G \typed A \subd A$,
% symetry: if $`G \typed A, B : s$ then $`G \typed A \subd B "<=>" `G \typed
% B \subd A$ and transitivity by \irule{SubTrans}.

@@

\begin{paragraph}{}
We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
%We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
%$T  `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
Our algorithmic judgement $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. 
% This notion is presented in \PVS{} by
% means of a $\mu_0$ operator which was readily adaptable in our setting
% (figure \vref{fig:mualgo-definition}). In Gang Chen's work, the notion is formalized by a
% new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator
% $\mualgo$, except ours finds $\Sigma_{lub}$s too.
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}
    \def\fCenter{\typea}
    \begin{center}
%     \WfEmpty\DP

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity and symmetry rules which are not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Admissibility of transitivity]
  If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
  \suba V : s$.
\end{theorem}
\begin{proof}
  By induction on the sum of the depths of the two derivations.
\end{proof}
\end{paragraph}

@@

Our interpretation will build a full-fledged \CIC term from its
algorithmic skeleton and a rich type written in \Russell. Obviously, we
cannot infer proof terms where they are needed in \CIC, but we can build
a \emph{partial} term, leaving typed holes where proofs are
needed. Hence we add a rule to build existential variables (or
metavariables) in the target calculus ($\typeq$ denotes the new system's typing judgement):

@@

The proof of this proposition is the main technical contribution of this
paper. It also represented the main difficulty in our work. 
The remainder of this section is organised as follows: first we will
define the explicit coercion derivation algorithm which will permit to
put proof obligations in the terms, then we will define the
interpretation of terms which is mutually recursive with the precedent
algorithm. Finally we will present a proof sketch of the aforementioned
proposition.

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations, hence we can derive a function from
the judgement which build a coercion given two types (we denote it $\coerce{`G}{T}{U}$). 
The side-conditions of the rules have not changed, so we omit them for
improved readability. The judgement $\subimpl{`G}{c}{T}{U}$ builds a
coercion $c$ from $T$ to $U$ in context $`G$, that is an object which
coerces an object of type $\ip{T}{`G}$ to an object of type $\ip{U}{`G}$
in context $\ipG{`G}$. We use the algorithmic types to drive the
derivation but the resulting object will be a well-typed \Coq{} term. We
could also directly use the interpreted types but that complicates the
proof.

@@

Coercions are formalised as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instantiation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(i.e., it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied.
Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgement $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and adapted for checking coherence 
of coercions, which we don't need here as we can get unicity
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respects some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this wasn't adapted to our setting, so we invented this
new presentation.

@@

We first prove reflexivity, symmetry and transitivity of the coercion
derivation algorithm, which extends previous proofs for the algorithmic
system with properties on
the generated coercions. Then we show substitutivity of the
interpretation: if $`G, x : U, `D \type t : T$ then $\ip{t[u/x]}{`G,
 `D[u/x]} = \ip{t}{`G, x : U, `D}[\ip{u}{`G}/x]$. We extend the notion
of coercion to contexts next and show stability of judgements under
context coercion. It is then possible to prove a lemma about commutativity of
substitutions with the interpretation and coercions. Then we get an
important corollary: if $`G \type u : U$, $\subimpl{`G}{c}{U}{V}{s}$ and
$`G, x : V \type T : s$ then $\ip{T[u/x]}{`G} `= \ip{T}{`G, x :
  V}[c[\ip{u}{`G}]/x]$. Essentially, it shows that coercions inserted
by the interpretation depend only on the context and hence could be
added later using substitution instead. This allows us to show that
equivalence is conserved by interpretation and finally that our
interpretation is correct.

\section{Implementation}
\label{sec:Program}
This mechanism has been implemented in \Coq and tested on simple
examples. We actually generate coercions simultaneously with typing 
in the implementation, which is sound thanks to the previous proofs. Our
typing is a clone of \Coq's original typing algorithm, so we benefit
from all the features of \Coq including implicit variables, notations and the existing
coercion system. The type-checker also transforms pattern-matching so
that an equality between the matched term and the pattern is present in
each branch typing context and it allows the user to put explicit holes in the term, for example:
\begin{code}
  Program Definition head \{ l : list A | l <> 0 \} : A :=
    match l with
    | hd :: tl -> hd
    | [] -> _
    end.
\end{code}
Here the second branch can't be taken because we would have $l = []$ and
$l \neq 0$, but we may not have any placeholder element of type $A$.

@@

\paragraph{Acknowledgements}
I'd like to thank Christine Paulin-Mohring for directing this work.
\bibliography{my,bib-joehurd,barras,pvs-bib,bcp,Luo,biblio-subset,%
cparent,demons,demons2,demons3}
\bibliographystyle{splncs}
% \renewcommand{\thefootnote}{}
% \footnotetext{This article was typeset in \LaTeX~using the
%   \texttt{Concrete Math} font}
